Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-09 paper/examples.tex
working on composition examples

Our algorithm splits \texttt{dot} into the two functions:
%
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()), (x2,y2,()))    = ((), (x1*x2)+(y1*y2))`
2`fun dot2 ((((),(),z1),((),(),z2)),l) = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, except
that they write the precomputation into a mutable cache.

@@

Our next example, \texttt{exp}, is a mainstay of the partial evaluation
literature (for example, in \cite{jones96}). \texttt{exp} recursively computes
$b^e$ using exponentiation by squaring, where $e$ is known at the first stage,
and $b$ is known at the second stage. 
%
\begin{lstlisting} 
1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`gr{`1`e`3`}`1` : ^int) = 
  if `3`gr{`1`e == 0`3`}`1` then 
    next{`2`literalone`1`}
  else if `3`gr{`1`(e mod 2) == 0`3`}`1` then
    exp(next{`2`b*b`1`},`3`gr{`1`e/literaltwo`3`}`1`)
  else 
    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`gr{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
\end{lstlisting}
%
Because \texttt{exp} is a recursive function whose conditionals test the parity
of the exponent argument, the sequence of branches taken corresponds exactly to
the binary representation of $e$. Partially evaluating \texttt{exp} with $e$
eliminates all of the conditionals, selecting and expanding the appropriate
branch in each case.

@@

1`fun exp1 (b : unit, e : int) =
  if e==0 then
    ((), Zero)
  else if (e mod 2)==0 then 
    ((), Even (#2 (exp1 ((), e/2))))
  else 
    ((), Odd (#2 (exp1 ((), (e-1)/2))))`

@@

This is an instance where stage-splitting is more practical than partial
evaluation; if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup by not expanding some calls to
\texttt{part}, would result in duplicating first-stage computations.)

@@

Note that the recursive \texttt{part} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{part}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.

To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
the second pass, evaluate \texttt{qs2} on that tree and each rank.
\begin{lstlisting}
1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
\end{lstlisting}
%

@@

Programs featuring staged structure are particularly important in
real-time computer graphics since modern graphics architectures
\emph{require} that graphics computations be organized into a pipeline
of stages which perform increasingly fine-grained computations (e.g.,
per object, per screen region, per pixel). Many computations in later
stages use the results from a single computation in an earlier
stage\,\cite{OpenGL4Spec}.

While performant, manually partitioning programs by stage results in
complex code where invariants must hold across different stages, and
local changes to the logic of one stage may require changes to that of
upstream stages other stages to ``plumb'' its inputs through the
pipeline.  Thus, manual separation of programs by stage breaks key
programming abstractions such as composition and modularity.  Graphics
researchers therefore have suggested using mixed-staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express logic related to
a single application concern conveniently as a single staged function
(referred to as a pipeline program)\,\cite{Foley:2011}.

Pipeline programs can be directly expressed as staged functions in
\lang\ with splitting used to generate monostaged code required for
each graphics pipeline stage. Composition of such functions in
\lang\ allows more complex pipelines to be expressed in a modular
programming style that is becoming increasingly desirable as graphics
programs increase in complexity\,\cite{Foley:2011,He:2014}.

A simple example of composition of staged functions (representing graphics
pipeline programs) is given below.  The code defines a pipeline as a
staged function that takes an object definition in the first stage
(\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second
stage and emits the color of the object at the specified pixel. The
code defines a pipeline combinator \texttt{shade} which modulates the
results of the object's reflectance at a given pixel with the results
of a function that computes the object's albedo (surface color).
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{  type coord = int * int
        type color = int        }`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline, albedo : pipeline) : pipeline =
  fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl (obj,next{2`xy`})`2`} * prev{`1`albedo (obj,next{2`xy`})`2`}`
\end{lstlisting}
%
Whereas prior systems provide combinators like \texttt{par} as
language features\,\cite{Foley:2011} to support composition, \lang\ is
powerful enough to define those combinators in-language.

>>>>

Our algorithm splits \texttt{dot} into the two functions:
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()), (x2,y2,()))    = ((), (x1*x2)+(y1*y2))`
2`fun dot2 ((((),(),z1),((),(),z2)),l) = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, except
that they write the precomputation into a mutable cache.

@@

Our next example, \texttt{exp}, is a mainstay of the partial evaluation
literature (for example, in \cite{jones96}). \texttt{exp} recursively computes
$b^e$ using exponentiation by squaring, where $e$ is known at the first stage,
and $b$ is known at the second stage. 
%
\begin{lstlisting} 
1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`gr{`1`e`3`}`1` : ^int) = 
 if      `3`gr{`1`e == 0`3`}`1`       then next{`2`literalone`1`}
 else if `3`gr{`1`e mod 2 == 0`3`}`1` then exp(next{`2`b*b`1`},`3`gr{`1`e/literaltwo`3`}`1`)
 else                     next{`2`b*prev{`1`exp(next{`2`b*b`1`},`3`gr{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
\end{lstlisting}
%
Because \texttt{exp} is a recursive function whose conditionals test the parity
of the exponent argument, the sequence of branches taken corresponds exactly to
the binary representation of $e$. Partially evaluating \texttt{exp} with $e$
eliminates all of the conditionals, selecting and expanding the appropriate
branch in each case.

@@

1`fun exp1 (b : unit, e : int) =
  if      e == 0       then ((), Zero)
  else if e mod 2 == 0 then ((), Even (#2 (exp1 ((), e/2))))
  else                      ((), Odd  (#2 (exp1 ((), (e-1)/2))))`

@@

This is an instance where stage-splitting is more practical than partial
evaluation; if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup, by not expanding some calls to
\texttt{part}, would result in duplicating first-stage computations.)

@@

Note that the recursive \texttt{part} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{part}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.
%

\subsection {Mixed Map Combinator}

[working on this section!]

In \ref{sec:exampleQS}, we showed the output of splitting quickselect
applied to one list and $m$ queries.  Here, we generalize to mapping
quickselect onto a list of queries. Note that we have to define a
datatype for integer lists at the second stage, which is given by
\texttt{list2}.
\begin{lstlisting} 
1`atsignnext{`2`
  datatype list2 = Empty2 | Cons2 of int * list2
`1`}

map : (^list * $`2`int`1` -> $`2`int`1`)
      -> ^list * $`2`list2`1` -> $`2`list2`1`
fun map f (l, q) = next {`2`
  let 
  fun m Empty2 = Empty2
    | m (Cons2(h,t)) =
                    Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
  in m prev{`1`q`2`}`1`}

val mapqss = map qss`
\end{lstlisting}
The \texttt{map} function splits into the following two functions:
\begin{lstlisting} 
2`datatype list2 = Empty2 | Cons2 of int * list2

`1`fun map1 f = (fn (l,()) => ((),#2 (f (l,()))),())
`2`fun map2 (f,()) ((l,q), p : tree) =
  let 
  fun m Empty2 = Empty2
    | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
  in m q

`1`val (mapqss1,()) = map1 qs1
`2`val mapqss2 = map2 (qs2, ())`
\end{lstlisting}
%
Indeed, \texttt{qs1}, which is passed to \texttt{map1}, is evaluated only in the
first pass, whereas \texttt{qs2}, which is passed to \texttt{map2}, is evaluated
in the second pass, once per query \texttt{k}.

@@

Composable staged programs are particularly important the domain of real-time graphics.
This need arises because modern graphics architectures actually
\emph{require} that graphics computations be structured as a pipeline
of stages which perform increasingly fine-grained computations (e.g.,
per-object, per-screen region, per-pixel). Many computations in later
stages use the results from a single computation in an earlier
stage\,\cite{OpenGL4Spec}.

The standard way to program these graphics pipelines is to define one program (usually called a \emph{shader}) per stage.
In other words, the programmer is expected to manually split their multi-stage programs.
Yet manually partitioning programs by stage results in
complex code where invariants must hold across different stages and
local changes to the logic of one stage may require changes to that of
upstream stages to communicate required inputs through the pipeline. 
This harms composition and modularity.  
Graphics researchers therefore have suggested using mixed-staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express graphics program logic 
for the entire pipeline as a single staged function\,\cite{Foley:2011}.

As a language, \lang\ is well suited for specification of such functions,
and we give a simple example below.
In it, we consider a graphics pipeline program to be a
staged function that takes an object definition in the first stage (\texttt{object})
and a pixel coordinate in the second stage ($\fut$\texttt{coord}), 
and emits the color of the object at the specified pixel ($\fut$\texttt{color}). 
Given two such multistage functions, we then define a combinator
that multiplies their results pointwise in the second 
stage.\footnote{Although this is a general pointwise multiplication combinator, 
the variable names suggest a possible interpretations of the inputs and output:
the first input function calculates the albedo of an object (a measure of how much light it reflects),
the second input calculates the object's incoming light,
and so the output (the product of these terms) is a function that calculates 
the outgoing light from an object.}
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{  type coord = int * int
        type color = ...        }`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline, albedo : pipeline) : pipeline =
  fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl (obj,next{2`xy`})`2`} * prev{`1`albedo (obj,next{2`xy`})`2`}`
\end{lstlisting}
%
A key strength of \lang\ is that it allows users to define combinators like this one as higher order functions.
Prior work\,\cite{Foley:2011} does not support higher order functions.

### 2016-01-09 paper/semantics.tex
working on composition examples

Multistage functions, such as \texttt{qss} from \ref{sec:overview},
can be represented as terms with a type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply such a function $f$ to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

Moreover, the reuse of first-stage computations across
multiple second-stage computations can even be encoded within \lang. 
The following program runs many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list:
\begin{lstlisting}
2`prev{`1` let val list = `3`gr{`1`[7,4,2,5,9,...,3]`3`}`1` in 
next{ `2`let fun qs k = prev{`1`qss (list,next{`2`k`1`})2`}`
      2`in (qs k1,...,qs km)`1`}`2`}`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let fun qs k = (let val y = ... in y)
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a large nest of case statements open on $k$.

>>>>

Multistage functions, such as \texttt{qss} from \ref{sec:overview},
can be represented as terms with a type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply such a function $f$ to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}
Moreover, the reuse of first-stage computations across
multiple second-stage computations can even be encoded within \lang. 
The following program runs many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list:
\begin{lstlisting}
2`prev{`1` let val list = `3`gr{`1`[7,4,2,5,9,...,3]`3`}`1` in 
next{ `2`let fun lookup k = prev{`1`qss (list,next{`2`k`1`})2`}`
      2`in (lookup k1,...,lookup km)`1`}`2`}`
\end{lstlisting}
Observe that this code evaluates \texttt{qss} only once
and subsitutes its result into the body of the function \texttt{lookup},
which is then called many times in the second stage.
